hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2025-01-07 inst/manuscript/paper.Rmd
writing discussion

talk about how these models also represent local adaptation and heritable plasticity

>>>>

With **hatchR** we present a software ecosystem that bridges the analytical gap of predicting developmental phenology for wild fishes and develops a formal framework for applying effective value models from user-provided parameterizations. To do so, the software is bundled in two forms, a fully customizable R package, especially useful for running repetitive analyses and a graphical-user-interface designed for ease of use and tasks that may only run once or a handful of times. Both applications allows users to import their data, run basic data checks, and apply historic model parameterizations for salmonids or create their own species- or population specific parameterization. Additionally, we provide substantial documentation on the **hatchR** website (<https://bmait101.github.io/hatchR/>) to walk users through basic to advanced applications of programmatic platform.

In the application of using effective value models **hatchR** and the user make some key assumptions. Particularly, numerous studies have indicated that stressful environmental conditions can cue fish to prematurely hatch or emerge from their environment. These include water quality like dissolved oxygen, pH, or salnity, pathogens, and even mechanical agitation (reviewed in @quinn2018 and @cowan2024). Moreover, while the **hatchR** provides point estimates for developmental phenology, spawning and developmental within populations of fishes both occur as distributions as opposed single events @mason1976 and we encourage users to either predict phenology using early, peak, and late frameworks (*e.g.*, 5th, 50th, and 95th quantiles) or with real or modeled distributions.

To date, the application of the effective value framework has largely focused using species-specific models to predict phenology in wild environments (see @adelfio2024, @austin2019). However, we emphasize that the statistical relationships at the basis of these models are fundamentally reaction norms, meaning that both the intercept and slope of the responses of fishes to temperature are reflective of family- or population-level genetic variation, genetic x environment responses, or may be indicative of phylogenetic differences among species @west-eberhard2003. For instance, while @sparks2017 did not find differences in the rates of development between the focal populations of their study, they did observe family-level genetic x environment interactions across different thermal regimes. Similarly, when they reparameterized their models using fish from western Alaska, they found differences in both the slope and intercept of the model relative to the original parameterization from @beacham1990, notably derived from multiple Canadian stocks. The western Alaskan fish generally developed more slowly than their southerly counterparts, consistent with cogradient variation @conover2009, @sparks2022. In this sense, we encourage users not to only think about the end points of these models (days to development) but also how the statistical relationships they are premised on inform our understanding of micro- and and macro-evolutionary processes.
